---
title: "Prelim_single_Cell"
author: "Kenneth Askelson"
date: "2025-08-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## PSC8 

Original file names looked like 'PSC8-DM-032725.scRNA.filtered.barcodes.tsv.gz', seurat doesn't like this so I need to remove 'PSC8-DM-032725.scRNA.filtered.' from each file.

```{r PSC8 load files}
library("Seurat")
library("ggplot2")
library("dplyr")
library("tidyr")

data_dir <- 'PSC8_Dragen'
list.files(data_dir) # Should show barcodes.tsv, genes.tsv, and matrix.mtx
expression_matrix <- Read10X(data.dir = data_dir)
seurat_object = CreateSeuratObject(counts = expression_matrix, min.features = 300, min.cells = 3)

seurat_object
```

Now we take a first pass at QC from the data 

```{r first pass QC PSC 8, echo=FALSE}
VlnPlot(seurat_object, features = c("nFeature_RNA", "nCount_RNA"), ncol = 2)

FeatureScatter(seurat_object, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")

```

```{r threshold checking and filtering}

#hist of nCount_RNA

ggplot(seurat_object@meta.data, aes(x = nCount_RNA)) +
  geom_histogram(bins = 100, fill = "lightgreen", color = "black") +
  geom_vline(aes(xintercept = 7000), colour="black") +
  labs(title = "Total Number of Counts (nCount_RNA)", x = "nCount_RNA", y = "Number of Cells")

#hist of nFeature_RNA

ggplot(seurat_object@meta.data, aes(x = nFeature_RNA)) +
  geom_histogram(bins = 100, fill = "red", color = "black") +
  geom_vline(aes(xintercept = 2500), colour="black") +
  labs(title = "Total Number of Counts (nFeature_RNA)", x = "nFeature_RNA", y = "Number of Cells")


seurat_object_filtered <- subset(seurat_object, subset = nFeature_RNA > 150 & nFeature_RNA < 2500 & nCount_RNA < 7000)


```
Repeat Violin and Scatter plot and normalize

```{r}
VlnPlot(seurat_object_filtered, features = c("nFeature_RNA", "nCount_RNA"), ncol = 2)

FeatureScatter(seurat_object_filtered, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")

seurat_object_filtered_normalized <- NormalizeData(seurat_object_filtered)

seurat_object_filtered_normalized
```
Identify variable genes

```{r}

seurat_object_filtered_normalized_selected <- FindVariableFeatures(seurat_object_filtered_normalized, selection.method = "vst", nfeatures = 2000)

top10 <- head(VariableFeatures(seurat_object_filtered_normalized_selected), 10)

# plot variable features with and without labels
plot1 <- VariableFeaturePlot(seurat_object_filtered_normalized_selected)
plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE)
plot2

```
Now we process data for PCA and identify informative # of PCs
```{r}
all.genes <- rownames(seurat_object_filtered_normalized_selected)

seurat_object_filtered_normalized_selected_scaled <- ScaleData(seurat_object_filtered_normalized_selected, features = all.genes)

seurat_object_filtered_normalized_selected_scaled_PCA <- RunPCA(seurat_object_filtered_normalized_selected_scaled, features = VariableFeatures(object = seurat_object_filtered_normalized_selected_scaled))

print(seurat_object_filtered_normalized_selected_scaled_PCA[["pca"]], dims = 1:5, nfeatures = 5)

VizDimLoadings(seurat_object_filtered_normalized_selected_scaled_PCA, dims = 1:2, reduction = "pca")

DimPlot(seurat_object_filtered_normalized_selected_scaled_PCA, reduction = "pca") + NoLegend()

DimHeatmap(seurat_object_filtered_normalized_selected_scaled_PCA, dims = 1:15, cells = 500, balanced = TRUE)

ElbowPlot(seurat_object_filtered_normalized_selected_scaled_PCA)
```
Looks like 6 for a cutoff? Now we try to cluster by cell types

```{r}
seurat_object_filtered_normalized_selected_scaled_PCA_cellcluster <- FindNeighbors(seurat_object_filtered_normalized_selected_scaled_PCA, dims = 1:6)

seurat_object_filtered_normalized_selected_scaled_PCA_cellcluster <- FindClusters(seurat_object_filtered_normalized_selected_scaled_PCA_cellcluster, resolution = 0.5)

seurat_object_filtered_normalized_selected_scaled_PCA_cellcluster <- RunUMAP(seurat_object_filtered_normalized_selected_scaled_PCA_cellcluster, dims = 1:6)

DimPlot(seurat_object_filtered_normalized_selected_scaled_PCA_cellcluster, reduction = "umap")

```
Now let's see if our genes of interested are expressed in any clusters. We will also find, for each cluster, every gene that is significantly differently expressed from other clusters.  

"gene:Ha412HOChr10g0435441", "gene:Ha412HOChr10g0435451", "gene:Ha412HOChr10g0435491", "gene:Ha412HOChr10g0435511"

Ha412HOChr10g0435621 wasn't found

```{r}
FeaturePlot(seurat_object_filtered_normalized_selected_scaled_PCA_cellcluster, features = c("gene:Ha412HOChr10g0435441", "gene:Ha412HOChr10g0435451", "gene:Ha412HOChr10g0435491", "gene:Ha412HOChr10g0435511"))

VlnPlot(seurat_object_filtered_normalized_selected_scaled_PCA_cellcluster, features = c("gene:Ha412HOChr10g0435441", "gene:Ha412HOChr10g0435451", "gene:Ha412HOChr10g0435491", "gene:Ha412HOChr10g0435511"))

seurat.markers <- FindAllMarkers(seurat_object_filtered_normalized_selected_scaled_PCA_cellcluster, only.pos = TRUE)

seurat.markers %>%
  group_by(cluster) %>%
  dplyr::filter(avg_log2FC > 1)

seurat.markers %>%
  group_by(cluster) %>%
  dplyr::filter(avg_log2FC > 1) %>%
  slice_head(n = 10) %>%
  ungroup() -> top10

DoHeatmap(seurat_object_filtered_normalized_selected_scaled_PCA_cellcluster, features = top10$gene)

```
Looks like something is going in with gene:Ha412HOChr10g0435511 in cluster 4, and cluster 4 more generally

## XRQ

Rinse and repeat

```{r}
data_dir_XRQ <- 'XRQ_Dragen'
list.files(data_dir) # Should show barcodes.tsv, genes.tsv, and matrix.mtx
expression_matrix_XRQ <- Read10X(data.dir = data_dir_XRQ)
seurat_object_XRQ = CreateSeuratObject(counts = expression_matrix_XRQ, min.features = 300, min.cells = 3)

seurat_object_XRQ
```
Way more cells (19,062)!!!

```{r}
VlnPlot(seurat_object_XRQ, features = c("nFeature_RNA", "nCount_RNA"), ncol = 2)

FeatureScatter(seurat_object_XRQ, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")

```
```{r}
#hist of nCount_RNA

ggplot(seurat_object_XRQ@meta.data, aes(x = nCount_RNA)) + xlim(0, 5000) +
  geom_histogram(bins = 100, fill = "lightgreen", color = "black") +
  geom_vline(aes(xintercept = 1000), colour="black") +
  labs(title = "XRQ Total Number of Counts (nCount_RNA)", x = "nCount_RNA", y = "Number of Cells")

#hist of nFeature_RNA

ggplot(seurat_object_XRQ@meta.data, aes(x = nFeature_RNA)) + 
  geom_histogram(bins = 100, fill = "red", color = "black") +
  geom_vline(aes(xintercept = 1000), colour="black") +
  labs(title = "XRQ Total Number of Counts (nFeature_RNA)", x = "nFeature_RNA", y = "Number of Cells")

seurat_object_XRQ_filtered <- subset(seurat_object_XRQ, subset = nFeature_RNA > 145 & nFeature_RNA < 1000 & nCount_RNA > 160 & nCount_RNA < 1000)

seurat_object_XRQ_filtered
```
```{r}
VlnPlot(seurat_object_XRQ_filtered, features = c("nFeature_RNA", "nCount_RNA"), ncol = 2)

FeatureScatter(seurat_object_XRQ_filtered, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")

```
```{r}
seurat_object_XRQ_filtered_normalized <- NormalizeData(seurat_object_XRQ_filtered)

#Feature selection

seurat_object_XRQ_filtered_normalized_selected <- FindVariableFeatures(seurat_object_XRQ_filtered_normalized, selection.method = "vst", nfeatures = 2000)

# Identify the 10 most highly variable genes
top10 <- head(VariableFeatures(seurat_object_XRQ_filtered_normalized_selected), 10)

# plot variable features with and without labels
plot1 <- VariableFeaturePlot(seurat_object_XRQ_filtered_normalized_selected)
plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE)
plot2

```
PCA and Elbow plot
```{r}
all.genes_XRQ <- rownames(seurat_object_XRQ_filtered_normalized_selected)

seurat_object_XRQ_filtered_normalized_selected_scaled <- ScaleData(seurat_object_XRQ_filtered_normalized_selected, features = all.genes_XRQ)

seurat_object_XRQ_filtered_normalized_selected_scaled_PCA <- RunPCA(seurat_object_XRQ_filtered_normalized_selected_scaled, features = VariableFeatures(object = seurat_object_XRQ_filtered_normalized_selected_scaled))

print(seurat_object_XRQ_filtered_normalized_selected_scaled_PCA[["pca"]], dims = 1:5, nfeatures = 5)

VizDimLoadings(seurat_object_XRQ_filtered_normalized_selected_scaled_PCA, dims = 1:2, reduction = "pca")

DimPlot(seurat_object_XRQ_filtered_normalized_selected_scaled_PCA, reduction = "pca") + NoLegend()

DimHeatmap(seurat_object_XRQ_filtered_normalized_selected_scaled_PCA, dims = 1:15, cells = 500, balanced = TRUE)

ElbowPlot(seurat_object_XRQ_filtered_normalized_selected_scaled_PCA)

```


```{r}
seurat_object_XRQ_filtered_normalized_selected_scaled_PCA_cellcluster <- FindNeighbors(seurat_object_XRQ_filtered_normalized_selected_scaled_PCA, dims = 1:10)

seurat_object_XRQ_filtered_normalized_selected_scaled_PCA_cellcluster <- FindClusters(seurat_object_XRQ_filtered_normalized_selected_scaled_PCA_cellcluster, resolution = 0.5)

seurat_object_XRQ_filtered_normalized_selected_scaled_PCA_cellcluster <- RunUMAP(seurat_object_XRQ_filtered_normalized_selected_scaled_PCA_cellcluster, dims = 1:10)

DimPlot(seurat_object_XRQ_filtered_normalized_selected_scaled_PCA_cellcluster, reduction = "umap")

FeaturePlot(seurat_object_XRQ_filtered_normalized_selected_scaled_PCA_cellcluster, features = c("gene:Ha412HOChr10g0435441", "gene:Ha412HOChr10g0435451", "gene:Ha412HOChr10g0435491", "gene:Ha412HOChr10g0435511"))

VlnPlot(seurat_object_XRQ_filtered_normalized_selected_scaled_PCA_cellcluster, features = c("gene:Ha412HOChr10g0435441", "gene:Ha412HOChr10g0435451", "gene:Ha412HOChr10g0435491", "gene:Ha412HOChr10g0435511"))


```
```{r}
seurat_XRQ.markers <- FindAllMarkers(seurat_object_XRQ_filtered_normalized_selected_scaled_PCA_cellcluster, only.pos = TRUE)

seurat_XRQ.markers %>%
  group_by(cluster) %>%
  dplyr::filter(avg_log2FC > 1)

seurat_XRQ.markers %>%
  group_by(cluster) %>%
  dplyr::filter(avg_log2FC > 1) %>%
  slice_head(n = 10) %>%
  ungroup() -> top10

DoHeatmap(seurat_object_XRQ_filtered_normalized_selected_scaled_PCA_cellcluster, features = top10$gene)

```



